Split-Apply-Combine with Dynamic Grouping

### 2023-01-02 paper/article.Rnw
writing

\usepackage{amsmath, amssymb}
\usepackage[ruled, linesnumbered]{algorithm2e}

@@

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Split-Apply-Combine with Collapsing Groups}
\Plaintitle{Split-Apply-Combine with Collapsing Groups}
\Shorttitle{Split-Apply-Combine with Collapsing Groups}

@@

@@

\section{The accumulate R package}\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. There are two main functions in \pkg{accumulate} that offer
slightly different interfaces.
\begin{verbatim}
  accumulate(data, collapse, test, fun, ...)
  cumulate(data, collapse, test, ...)
\end{verbatim}
Here \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence (as a \class{formula} or a \class{data frame}), and \texttt{test} is a
function that accepts a subset of \texttt{data} and returns a boolean that
indicates whether a subset is suited for aggregation. In \code{accumulate()},
the parameter \code{fun} represents an aggregation function that is applied to
every column of \code{data}, and the ellepsis (\code{...}) is for arguments
that are passed as extra argument to \code{fun}. The interface of
\code{accumulate()} is modeled after the \code{aggregate()} function in
\proglang{R}.  In \code{cumulate}, the ellepsis is a sequence of
comma-separated \code{name=expression} pairs in the style of \code{summarise()}
from the \code{dplyr} package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
It is not uncommon for users to use multiple variables for defining collapsing
schemes. For example when grouping by two categorical variables $A$ and
$B$, one may first want to collapse variable $B$ into a
variable $B1$ and if that is not sufficient, grouping
in the second variable is dropped so grouping is only defined by $A$.
In \code{accumulate()} such a scheme is denoted with the formula \code{A*B \~{}
A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
that is attempted first, and the right-hand-side denotes consecutive collapse
levels separated by \code{+}.

The following tables illustrate the idea. The left table represents the data
set to be aggregated by $A$ and $B$. In this example we wish to
compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
least three records in a group, otherwise we collapse one step (we set values
of $Y$ to unique powers of two, so it is clear which records are used
to compute each value of $t_Y$).
%
\begin{center}
\begin{tabular}{cccl}
\multicolumn{4}{l}{Input data}\\
\hline
$A$ & $B$ & $B1$ & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{Output}\\
\hline
$A\times B$ & \code{level} & $t_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 7  \\
\rowcolor{green!25} 2 12        & 1             & 56 \\
\rowcolor{green!25} 2 13        & 1             & 56 \\
\rowcolor{blue!25}  3 21        & 2             & 448\\
\rowcolor{blue!25}  3 22        & 2             & 448\\
\rowcolor{blue!25}  3 12        & 2             & 448\\
\hline
\end{tabular}
\end{center}
%
The table on the right represents the output. Colors indicate which data was
used. The First row in the output represents group
$(A=1,B=11)$. The collapsing level is zero, which means
that no collapsing was necessary. Indeed, in the data table we see that there
are three rows with $A=1$ and $B==11$ with $Y$ values $1, 2$,
and $7$, resulting in $t_Y=7$ for this group.

Next, the algorithm tries to compute the total for group
$(A=2,B=12)$ (in green) but finds that there are only two
such rows. It now defines a new group $(A=2,B1=1)$ and finds
that there are three records in that group so it computes $t_Y=8+16+32=56$.
Similarly, it finds one record with $(A=2,B=13)$ and
collapses to $(A=2,B1=1)$, again yielding $t_y=56$.

Finally, for $(A=2,B=21)$ the algorithm finds only a single
record. It collapses to $(A=2,B1=2)$ and finds two so it
collapses further to $(A=2)$ to find three records.  This yields
$t_Y=64+128+256=448$. Similarly, the groups $(A=3, B=22)$ and
$(A=3,B=12)$ are collapsed to $(A=2)$.

In R code the previous example can be executed as follows.
<<>>=
library(accumulate)
input <- data.frame(
     A  = c(1,1,1,2,2,2,3,3,3)
   , B  = c(11,11,11,12,12,13,21,22,12)
   , B1 = c(1,1,1,1,1,1,2,2,1)
   , Y  = 2^(0:8)
)

cumulate(input, collapse = A*B ~ A*B1 + A
        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
@
%
The formula interface allows users to quickly experiment with different
collapsing schemes, but only after the necessary variables have been added to
the original data frame.

In order to use the data frame interface, the input dataset needs to have the
most fine-grained grouping variable. In the running example this is
$A\times B$, so we need to combine that into a single variable, and remove
the other ones.
<<>>=
input$AB <- paste(input$A, input$B, sep="-")
input <- input[-(1:3)]
input
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
@
In this data frame, two consecutive columns should be read as a child-parent
relation. For example, in the first collapsing step the groups defined by
\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
In this artificial example the codes do not mean anything, but in realistic
cases where codes represent a (hierarchical) classification, domain experts
usually have a good grasp of which codes can be combined.

\subsection{Balanced and unbalanced hierarchical classifications}

\subsection{Specifying tests}

\subsection{Programming over accumulate }

\section{Extensive example: imputing economic data} \label{sect:example}

>>>>

\usepackage{amsmath, amssymb}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{tikz}

@@

% This allows referencing institutional authors. 
% See: https://tex.stackexchange.com/questions/162659/formatting-of-institution-as-author-with-natbib-and-numbered-references
\newcommand{\xfnm}[1][]{\ifx!#1!\else\unskip,\space#1\fi}

@@

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Split-Apply-Combine with Dynamic Grouping}
\Plaintitle{Split-Apply-Combine with Dynamic Grouping}
\Shorttitle{Split-Apply-Combine with Dynamic Grouping}

@@

\section{Dynamic grouping} \label{sect:dynamicgrouping}
In this section the illustrate the concept of dynamic grouping with a minimal
worked example. This example is not very realistic, but it is constructed
to be simple enough so the whole procedure can be followed in detail.

We consider a data set with three categorical variables $A$, $B$ and $B_1$, and
one numerical variable $Y$.  Variable $A$ has levels $\{1, 2,3\}$ and variable
$B$ is a hierarchical classification with levels $\{11,12,13,21,22\}$. Variable
$B_1$ is a coursegraining of $B$: for each record the value for $B_1$ is the
first digit of $B$. Hence, $B_1$ has levels $\{1,2\}$.

Our goal is to compute the sum $t_Y$ over $Y$, grouped by $A\times B$.  We
impose the condition that there must be at least three records in each group.
If a certain group $(a\in A,b\in B)$ has less than three records, we attempt
to compute the value for that group over records in $(a,b_1)$ where $b_1$ is
obtained by taking the first digit of $b$. If we then still have less than
three records, we take records of group $a$ to determine the value for $(a,b)$.

The following tables illustrate the idea. The left table represents the data
set to be aggregated by $A$ and $B$.  The table on the right represents the
output. Colors indicate which data was used. 
%
\begin{center}
\begin{tabular}{cccl}
\multicolumn{4}{l}{Input data}\\
\hline
$A$ & $B$ & $B1$ & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{Output}\\
\hline
$A\times B$ & \code{level} & $t_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 7  \\
\rowcolor{green!25} 2 12        & 1             & 56 \\
\rowcolor{green!25} 2 13        & 1             & 56 \\
\rowcolor{blue!25}  3 21        & 2             & 448\\
\rowcolor{blue!25}  3 22        & 2             & 448\\
\rowcolor{blue!25}  3 12        & 2             & 448\\
\hline
\end{tabular}
\label{eq:example}
\end{center}
%
The First row in the output represents group $(A=1,B=11)$. The collapsing level
is zero, which means that no collapsing was necessary. Indeed, in the data
table we see that there are three rows with $A=1$ and $B==11$ with $Y$ values
$1, 2$, and $7$, resulting in $t_Y=7$ for this group.

Next, we try to compute the total for group $(A=2,B=12)$ (in green) but find
that there are only two such rows. We now define a new group $(A=2,B_1=1)$ and
find that there are three records in that group so we get 
$t_Y=8+16+32=56$.  Similarly, there is only one record with $(A=2,B=13)$. 
Collapsing groups to $(A=2,B1=1)$, yieds again $t_y=56$.

Finally, for $(A=2,B=21)$ there is only a single record. Collapsing to
$(A=2,B1=2)$ yields only two records, so we need to collapse further to $(A=2)$
and finally obtain three records.  This yields $t_Y=64+128+256=448$. Similarly,
the groups $(A=3, B=22)$ and $(A=3,B=12)$ are collapsed to $(A=2)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5%%%%%%%
\section{The accumulate R package}\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a fair amount of specification by the user. Besides the data to be
aggregated, one needs to specify the method(s) of aggregation, the collapsing
scheme, and the condition to decide whether a subset is fit for aggregation or
a next collapse is necessary. There are two main functions in \pkg{accumulate}
that offer slightly different interfaces.
\begin{verbatim}
  accumulate(data, collapse, test, fun, ...)
  cumulate(data, collapse, test, ...)
\end{verbatim}
Here \texttt{data} is a data frame holding data to be aggregated;
\texttt{collapse} represents the collapse sequence (as a \class{formula} or a
\class{data frame}), and \texttt{test} is a function that accepts a subset of
\texttt{data} and returns a boolean that indicates whether a subset is suited
for aggregation or not. In \code{accumulate()}, the parameter \code{fun}
represents an aggregation function that is applied to every column of
\code{data}, and the ellepsis (\code{...}) is for arguments that are passed as
extra argument to \code{fun}. The interface of \code{accumulate()} is modeled
after the \code{aggregate()} function in \proglang{R}.  In \code{cumulate}, the
ellepsis is a sequence of comma-separated \code{name=expression} pairs in the
style of \code{summarise()} from the \code{dplyr} package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
the formula interface.
<<>>=
library(accumulate)
input <- data.frame(
     A  = c(1,1,1,2,2,2,3,3,3)
   , B  = c(11,11,11,12,12,13,21,22,12)
   , B1 = c(1,1,1,1,1,1,2,2,1)
   , Y  = 2^(0:8)
)

cumulate(input, collapse = A*B ~ A*B1 + A
        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
@
%
Consider the formula \code{A*B ~ A*B1 + A} in the call to \code{cumulate()}.
The left-hand-side \code{A*B} is the target output grouping. The
ritght-hand-side is to be interpreted as the collapsing sequnce: if an instance
of \code{A*B} does not pass the test, then collapse to \code{A*B1}, and if that
does not pass the test collapse to \code{A}. If this final grouping also does
not pass the test, the result is \code{NA}.

Summarizing, the formula interface is always of the follwing form.
\begin{verbatim}
  Target grouping ~ Alternative1 + Alternative2 + ... + AlternativeN
\end{verbatim}

The formula interface allows users to quickly experiment with different
collapsing schemes. It does require that all categorical variables have been
added to the data. As an alternative, one can use the data frame specification,
which allows for a further separation between defining the collapsing scheme
and the actual data processing.

In order to use the data frame interface, the input dataset must include the
most fine-grained grouping variable. In the running example this is $A\times
B$, so we need to combine that into a single variable, and remove the other
ones.
<<>>=
input$AB <- paste(input$A, input$B, sep="-")
input <- input[-(1:3)]
input
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
csh
@
In this data frame, two consecutive columns should be read as a child-parent
relation. For example, in the first collapsing step the groups defined by
\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
In this artificial example the codes do not mean anything, but in realistic
cases where codes represent a (hierarchical) classification, domain experts
usually have a good grasp of which codes can be combined.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Balanced and unbalanced hierarchical classifications}
Hierarchical classifications are abundant in (official) statistics.
They represent a classification of entities into non-overlapping, nested
groupings. Examples include the International Standard Industrial Classification
of Economic Activities (ISIC, \citet{un2022isic}), the related 
Statistical Classification of Economic Activities in Europe (NACE, \citet{eu2006nace})
and the Euroean Skills, Competences, Qualifications and Occupations classification
(ESCO, \citet{eu2022esco}).

Hierarchical classifications offer a natural mechanism for collapsing fine-graind groupings
into larger groups because of the parent-child relationships. As an example consider
a small piece of the NACE classification.

\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=40mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
%  every node/.style = {shape=rectangle, rounded corners,
%    draw, align=center,
%    top color=white, bottom color=blue!20}]]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}}
      child {node {\code{0112}}} 
      child {node {\code{0113}}} }
    child { node {\code{012}}
      child {node {\code{0121}}}
      child {node {\code{0122}}} 
      child {node {\code{0123}}} 
      child {node {\code{0124}}} };
\end{tikzpicture}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Specifying tests}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Programming over accumulate }

\section{Extensive example: economic data} \label{sect:example}

@@
